---
title: "Data Analysis"
author: "Kaitlyn J. Tonra & Christopher D. Wells"
date: "1/11/2021"
output: html_document
---

This script is for calculating electivity and mortality for rubble and container level analyses. 

Setup chunk: Includes packages, plot style, and definitions for important variables 
```{r setup, include=FALSE}

knitr::opts_knit$set(root.dir = "../git/homomalla_settlement_choice")

library(boot) #load for bootstrapping
library(vegan) #load for PERMANOVA
library(bcaboot)

library(tidyverse)
library(grid)
library(gridExtra)
library(ggpubr)

rm(list =  ls())
graphics.off()

# Create a bootstrapping function
samplemean <- function(x, d) {return(mean(x[d]))} 

# Define styles for making plots later
electivity_plot <- list(
  ylim(-1,1),
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5))

electivity_theme <- function() {
  theme_minimal() +   theme(axis.title = element_text (size = 10),
                      axis.text = element_text (size = 10, color = "black"),
                      panel.background = element_blank (),
                      panel.grid.major = element_blank (),
                      panel.grid.minor = element_blank (),
                      plot.background = element_blank (),
                      axis.line = element_line(linewidth = 0.5, linetype = "solid", color = "black"),
                      axis.ticks = element_line(linewidth = 0.5, linetype = "solid", color = "black"))
  }

# Setup for calculating electivities (define parameters)
bootstrap <- 10000 #number of bootstraps
permutations <- 99999 #number of permutations
rubble_types <- 3 #number of different rubble types in the experiment
substratum_types <- 4 #number of substratum types from coralnet
settlement_day <- 6 #day chosen for the settlement electivity calculations
mortality_start <- 12 #first day to fit mortality rate to
mortality_end <- 26 #last day to fit mortality rate to

singles <- c(1, 4, 6)


custom_labels <- c(
  "bare" = "Bare",
  "cca" = "CCA",
  "ram" = "Ramicrusta",
  "spo" = "Sponge",
  "spon" = "Sponge"
)

custom_colors <- c(
  "bare" = "#ECC498", 
  "cca" = "#DC7177", 
  "ram" = "#C4501b", 
  "spo" = "#548E91",
  "spon" = "#548E91")
```


Import the data and rename. These are pulling data that use the second method of area calculations. 
```{r import}

#number of corals on each substratum type each day (raw data)
settlement_counts <- read_csv("../data/counts_raw.csv",
                              col_types = cols(container = col_character())) %>%
  filter(!container %in% singles)


#calculated area of each rubble (from data prep). Called rubble_area2 to correspond with method2
rubble_area <- read_csv("../code/rubble_area2.csv",
                        col_types = cols(container = col_character(),
                                         cca = col_double(),
                                         bare = col_double(),
                                         ram = col_double())) 

#calculated area of each substratum type in a container (from data prep). Called patch_type_area2 to correspond with method2. 
patch_type_area <- read_csv("../code/patch_type_area2.csv",
                            col_types = cols(container = col_double(),
                                             bare = col_double(),
                                             cca = col_double(),
                                             ram = col_double(),
                                             spon = col_double())) 
```


Isolate day 6 settlement counts so we can use it later for electivity calculation. This includes data for both rubble level and patch level analyses.
```{r isolate settlement day}

day6_settlement_counts <- settlement_counts %>%
  
# Add up settlers on each substratum
  mutate(bare = bt + bs + bb,
         cca = ct + cs + cb,
         ram = rt + rs + rb,
         spon = ot + os + ob) %>% 
  
# Isolate settlement day data  
  filter(day == settlement_day) %>% 

# Remove extra columns  
  select(-position:-ob) 

# View
day6_settlement_counts

# Needed this for just a sec
write.csv(day6_settlement_counts, "settlement.csv")
```

Arrange the data so it's in the right shape for calculating electivities on the rubble scale. 
```{r}

rubble_electivity_df <- day6_settlement_counts %>%
  
# Add a column to get total settlers in each container   
     mutate(total = bare + cca + ram + spon) %>% 
  
# This just changes the class type so we can use it later
     group_by(container, rubble_type)  %>% 
                  
# Arrange data organized by container and rubble type  
     summarize(total = sum(total)) %>% 
  
# Change the shape of the dataframe                 
     pivot_wider(names_from = rubble_type, values_from = total) %>% 
  
# Add the rubble_area dataframe so we can use it for electivities.                  
     merge(rubble_area, 
           by = "container", # Assign the correct area to each container
           suffixes = c(".count", ".areaprop"))# Label resulting columns that represent counts and proportion areas

rubble_electivity_df
                        
```


Arrange the data so it's in the right shape for calculating electivities on the patch (container) scale. 
```{r}

patch_electivity_df <- day6_settlement_counts %>%
  
# Add a column to get total settlers in each container   
   #  mutate(day6_settlement_counts, total = bare + cca + ram + spon) %>% 
  

# Change the shape of the dataframe                 
     pivot_longer(bare:spon, names_to = "patch_type", values_to = "num_individuals") %>% 

# This just changes the class type so we can use it later
     group_by(container, patch_type) %>% 
                  
# Arrange data organized by container and rubble type  
     summarize(total = sum(num_individuals), .groups = "drop") %>% 
    
# Flip wide again
     pivot_wider(names_from = patch_type, values_from = total) %>%
  
# Add the rubble_area dataframe so we can use it for electivities.                  
     merge(patch_type_area, 
           by = "container", # Assign the correct area to each container
           suffixes = c(".count", ".areaprop")) # Label resulting columns that represent counts and proportion areas
    
  
patch_electivity_df
                        
```


This chunk creates a function to calculate electivity based on the values assigned at the top of the Rmd. 
```{r Create function to calculate electivity}

get_electivity <- function(counts, areas){
  
  densities <- counts / areas

  total_density <- sum(densities)
  
  proportions <- densities / total_density    #i.e. Chesson's alpha
  
  ntypes <- length(counts)
  
  electivities <- (proportions - 1 / ntypes) / (proportions + 1 / ntypes)
  
  return(electivities)
  
}

```


Time to apply that function! This one is for Rubble level electivities. 
```{r Apply get_electivity}

elec <- data.frame("cca", "bare", "ram")

for(i in 1:6){
  counts <- c(rubble_electivity_df$cca.count[i], rubble_electivity_df$bare.count[i], rubble_electivity_df$ram.count[i])
  areas <- c(rubble_electivity_df$cca.areaprop[i], rubble_electivity_df$bare.areaprop[i], rubble_electivity_df$ram.areaprop[i])
  elec[i, ] <- get_electivity(counts, areas)
}

elec.rub <- elec %>%
  rename(cca = X.cca., bare = X.bare., ram = X.ram.) %>%
  pivot_longer(cca:ram, names_to = "rubbletype", values_to = "electivity") %>%
  filter(!is.na(electivity)) %>%
  mutate(electivity = as.numeric(electivity),
         rubbletype = as.factor(rubbletype))

elec.rub

```

Time to apply that function! This one is for container (patch) level electivities. :)
```{r Apply get_electivity}

elec <- data.frame("cca", "bare", "ram", "spon")

for(i in 1:6){
  counts <- c(patch_electivity_df$cca.count[i], 
              patch_electivity_df$bare.count[i], 
              patch_electivity_df$ram.count[i], 
              patch_electivity_df$spon.count[i])
  
  areas <- c(patch_electivity_df$cca.areaprop[i], 
             patch_electivity_df$bare.areaprop[i], 
             patch_electivity_df$ram.areaprop[i], 
             patch_electivity_df$spon.areaprop[i])
  
  elec[i, ] <- get_electivity(counts, areas)
}

elec

  elec.patch <- elec %>%
  rename(cca = X.cca., bare = X.bare., ram = X.ram., spon = X.spon.) %>%
  pivot_longer(cca:spon, names_to = "patchtype", values_to = "electivity") %>%
  filter(!is.na(electivity)) %>%
  mutate(electivity = as.numeric(electivity),
         patchtype = as.factor(patchtype))

elec.patch

```

Stats and graphs for rubble type electivity 
```{r, fig.width = 3, fig.height = 3}

ggdensity(elec.rub$electivity)
ggqqplot(elec.rub$electivity)
shapiro.test(scale(elec.rub$electivity)) #not normal

kruskal.test(electivity ~ rubbletype, data = elec.rub) #kruskal-wallis
pairwise.wilcox.test(elec.rub$electivity, elec.rub$rubbletype, p.adjust.method = "BH") #pairwise comparisons

cca <- bcajack(elec.rub$electivity[elec.rub$rubbletype == "cca"], B = 10000, func = samplemean, verbose = FALSE)
bar <- bcajack(elec.rub$electivity[elec.rub$rubbletype == "bare"], B = 10000, func = samplemean, verbose = FALSE)
ram <- bcajack(elec.rub$electivity[elec.rub$rubbletype == "ram"], B = 10000, func = samplemean, verbose = FALSE)

conf <- data.frame("rubbletype" = c("cca", "bare", "ram"),
                   "median" = c(cca$B.mean[2], bar$B.mean[2], ram$B.mean[2]),
                   "lwr" = c(cca$lims[1], bar$lims[1], ram$lims[1]),
                   "upr" = c(cca$lims[9], bar$lims[9], ram$lims[9]))

ggplot() +
  geom_errorbar(aes(x = rubbletype, ymin = lwr, ymax = upr), conf, width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(x = rubbletype, y = median, fill = rubbletype), conf, shape = 21, color = "black", size = 3) +
  scale_x_discrete(labels = c("Bare", "CCA", "Ramicrusta")) +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(x = "Rubble type", y = "Electivity index") +
  electivity_theme() #medians with 95% bias-corrected and accelerated bootstrap confidence intervals

```

Stats and graphs for container /patch type electivity 

```{r, fig.width = 3, fig.height = 3}
ggdensity(elec.patch$electivity)
ggqqplot(elec.patch$electivity)
shapiro.test(scale(elec.patch$electivity)) #not normal

kruskal.test(electivity ~ patchtype, data = elec.patch) #kruskal-wallis
pairwise.wilcox.test(elec.patch$electivity, elec.patch$patchtype, p.adjust.method = "BH") #pairwise comparisons

cca <- bcajack(elec.patch$electivity[elec.patch$patchtype == "cca"], B = 10000, func = samplemean, verbose = FALSE)
bar <- bcajack(elec.patch$electivity[elec.patch$patchtype == "bare"], B = 10000, func = samplemean, verbose = FALSE)
ram <- bcajack(elec.patch$electivity[elec.patch$patchtype == "ram"], B = 10000, func = samplemean, verbose = FALSE)
spo <- bcajack(elec.patch$electivity[elec.patch$patchtype == "spon"], B = 10000, func = samplemean, verbose = FALSE)

conf <- data.frame("patchtype" = c("cca", "bare", "ram", "spo"),
                   "median" = c(cca$B.mean[2], bar$B.mean[2], ram$B.mean[2], spo$B.mean[2]),
                   "lwr" = c(cca$lims[1], bar$lims[1], ram$lims[1], spo$lims[1]),
                   "upr" = c(cca$lims[9], bar$lims[9], ram$lims[9], spo$lims[9]))
```

Electivity fig!
```{r, fig.width = 3, fig.height = 3}
ggplot() +
  geom_errorbar(aes(x = patchtype, ymin = lwr, ymax = upr), conf, width = 0.3) + #medians with 95% bias-corrected and accelerated bootstrap confidence intervals
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(x = patchtype, y = median, fill = patchtype), conf, shape = 21, color = "black", size = 4) + 
  
  scale_x_discrete(labels = custom_labels) +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(x = "Substratum type", y = "Electivity index (E*)") +
  scale_fill_manual(values = custom_colors) +
  electivity_theme() + 
  theme(legend.position = "none")

ggsave("patch_electivity_medians.tiff", dpi = 600)

```


Counts data and prep for all count figs
```{r}
long_counts <- settlement_counts %>%
  
# Add up settlers on each substratum type
  mutate(bare = bt + bs + bb,
         cca = ct + cs + cb,
         ram = rt + rs + rb,
         spon = ot + os + ob) %>% 
  
# remove extra cols  
  select(day, container, bare:spon) %>%
  
# Make longer (add cols for patch and counts)
  pivot_longer(cols = c(bare, cca, ram, spon), names_to = "patch", values_to = "polyps")



alltime_counts <- long_counts %>%
  group_by(patch, day) %>%
# Calculate mean polyps per patch type per day with CIs
    summarize(
    sum_polyps = sum(polyps),
    avg_polyps = mean(polyps),  # Average number of polyps
    se_polyps = sd(polyps) / sqrt(n()),  # Standard error
    ci_lower = avg_polyps - 1.96 * se_polyps,  # Lower bound of 95% CI
    ci_upper = avg_polyps + 1.96 * se_polyps,  # Upper bound of 95% CI
    .groups = "drop"
  )
```


All polyps over time
```{r}
ggplot(alltime_counts, aes(x = day, y = sum_polyps, color = patch, group = patch)) +
  
  geom_line(size = 1) +  
  geom_point(size = 3) +  
  labs(
    x = "Day",
    y = "Number of polyps",
    title = "Total Polyps per Substratum Type",
    color = "Patch Type",
    fill = "Patch Type"
  ) +
  
  scale_color_manual(values = custom_colors, 
                     labels = custom_labels) + 
  
  scale_x_continuous(breaks = unique(alltime_counts$day)) +

  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16)
  ) +
  
  electivity_theme()

ggsave("total_polyps_all_days.tiff", dpi = 600, height = 6, width = 10)
```

Mean polyps over time with 95% CIs
```{r}
ggplot(alltime_counts, aes(x = day, y = avg_polyps, color = patch, group = patch)) +
  
# 95% CI ribbons, line, points
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = patch), alpha = 0.2) +  
  geom_line(size = 1) +  
  geom_point(size = 3) + 
  
  labs(
    x = "Day",
    y = "Mean number of polyps",
    title = "Mean Polyps per Substratum Type",
    color = "Patch Type",
    fill = "Patch Type"
  ) +
 
# Use custom colors for lines and fill
  scale_color_manual(labels = custom_labels, values = custom_colors) +  
  scale_fill_manual(labels = custom_labels, values = custom_colors) +  

# Use default colors
#  scale_color_discrete(labels = custom_labels) +  
#  scale_fill_discrete(labels = custom_labels) + 
  
# Days as X breaks  
  scale_x_continuous(breaks = unique(alltime_counts$day)) +  
  
  theme_minimal() +  
  theme(
    legend.position = "right",  
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),  
    plot.title = element_text(size = 16)  
  ) +
  
  electivity_theme()  



ggsave("mean_polyps_all_days1.tiff", dpi = 600, height = 6, width = 10)
```

Mortality props
```{r}
mortality_props <- alltime_counts %>%
  
  select(day, patch, sum_polyps, avg_polyps) %>%
  filter(day > 3) %>%
  filter(day < 17) %>%
  arrange(patch, day)%>%
  group_by(patch) %>%

  mutate(prop_remaining = case_when(day == 6 ~ 1,  # Day 6 is first day
    TRUE ~ avg_polyps / lag(avg_polyps)  # Otherwise calculate the ratio of avg_polyps to the lag of avg_polyps
  ))
```

Polyp mortality after Day 6
```{r}
ggplot(mortality_props, aes(x = day, y = prop_remaining, color = patch, group = patch)) +
  geom_line(size = 1) +  
  geom_point(size = 3) +  
  
  labs(
    x = "Day",
    y = "Proportion of Settled Polyps Remaining",
    title = "Proportion of surviving polyps",
    color = "Patch Type",
    fill = "Patch Type"
  ) +
  
    scale_color_manual(values = custom_colors, 
                     labels = custom_labels) +
  
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16)
  )

ggsave("10day_mortality.tiff", dpi = 600)
```
For binomial:
```{r}
binom_data <- long_counts %>%
  
  filter(day > 3) %>%
  filter(day < 17) %>%
  arrange(container, day, patch)




#%>%
  group_by(patch) %>%

  mutate(prop_remaining = case_when(day == 6 ~ 1,  # Day 6 is first day
    TRUE ~ polyps / lag(polyps)  # Otherwise calculate the ratio of avg_polyps to the lag of avg_polyps
  ))
```
```







```{r Calculate electivity (old)}
# # Calculate proportion of total settlers on each substratum
# rubble_merge_density <- mutate(rubble_merge, 
#                              density_bare = bare.x/bare.y,
#                              density_cca = cca.x/cca.y,
#                              density_ram = ram.x/ram.y)  
# 
# # Add Chesson's alphas
# rubble_alphas <- mutate(rubble_merge_density, bare_a = density_bare / (density_bare + density_cca + density_ram),
#                              cca_a = density_cca / (density_bare + density_cca + density_ram),
#                              ram_a = density_ram / (density_bare + density_cca + density_ram))
# rubble_alphas
# 
# # Calculate the electivities
# rubble_electivities <- mutate(rubble_alphas, bare_e = (bare_a - 1 / rubble_types) / (bare_a + 1 / rubble_types),
#                                              cca_e = (cca_a - 1 / rubble_types) / (cca_a + 1 / rubble_types),
#                                              ram_e = (ram_a - 1 / rubble_types) / (ram_a + 1 / rubble_types)) 
# 
# rubble_electivities
```


