---
title: "Plexaura homomalla settlement data prep"
author: "Kaitlyn J. Tonra & Christopher D. Wells"
date: "11/29/2021"
output: html_document
---
```{r setup, include = FALSE}
rm(list =  ls())
knitr::opts_knit$set(root.dir = "../git/homomalla_settlement_choice")
library(tidyverse)
library(dplyr)
```

# Data prep for P. homomalla settlement choice experiment from 2019. 

##This R markdown file pulls in raw CoralNet annotations and raw size measurements for rubble pieces and prepares them to be exported and used in calculating percent cover, electivity, mortality, and other analyses. 



1. Import: This chunk grabs CSV raw files and assigns appropriate data types to each column. The '_prep' suffix on file names denotes files that are being imported and prepared for analyses here. 
```{r import}

# coralnet_raw contains photo name, rubble type, side (top/bottom), container number, position of the annotation (row/col), and annotation (label).
# These same columns are imported and stored in coralnet_prep as appropriate column types. 

coralnet_prep <- read_csv("../data/coralnet_raw.csv", 
                 col_types = cols(name = col_character(),
                                  rubble_type = col_character(),
                                  side = col_character(),
                                  container = col_character(),
                                  row = col_double(),
                                  column = col_double(),
                                  label = col_character()))

# size_raw contains, for each rubble piece: Container number and rubble type
#                                           4 measurements of height at 4 different points 
#                                           3 measurements of top and bottom perimeter
#                                           3 measurements of top and bottom area
# The same colummns are stored in size_prep as appropriate column types. 

size_prep <- read_csv("../data/size_raw.csv",
                 col_types = cols(container = col_character(),
                                  rubble_type = col_character(),
                                  hgt1 = col_double(), 
                                  hgt2 = col_double(),
                                  hgt3 = col_double(),
                                  hgt4 = col_double(),
                                  perim_bot1 = col_double(),
                                  perim_bot2 = col_double(),
                                  perim_bot3 = col_double(),
                                  perim_top1 = col_double(),
                                  perim_top2 = col_double(),
                                  perim_top3 = col_double(),
                                  area_bot1 = col_double(),
                                  area_bot2 = col_double(),
                                  area_bot3 = col_double(),
                                  area_top1 = col_double(),
                                  area_top2 = col_double(),
                                  area_top3 = col_double())) 
```


------------ PART 1: CORALNET ANNOTATIONS FOR CALCULATING PROPORTIONS OF RUBBLE AND CONTAINER THAT ARE BARE/CCA/RAM/SPON AND T/S/B ------------------



2. Proportion of rubble surfaces: This uses data from the imported coralnet_prep to output the proportion of each rubble piece that is covered by each type of substratum type. This is done by calculating the number of annotations for each category and dividing these by the total number of annotations for each rubble. 
At the end of this chunk, 'proportioncover' has an entry for every rubble piece. Each row (rubble piece) has a column that gives the proportion of all annotations for that rubble that are considered: top, side, bottom, CCA, BARE, RAM, and OTHER. 
This final dataframe will not include raw numbers of annotations for each category (e.g., bare_top, ram_side), only proportions of the total. Top/side/bottom distinctions are not used in relationship to settlement, only for estimating total areas of the rubble, which is why we are separating side and substratum data. 
Variables with the prefix 'count_' are numbers of annotations and variables named with 'prop_'are proportions of the total number. Each rubble piece has <800 annotations total but they vary because of the photo annotation software that was used; each photo had random points with some located off-rubble. 

```{r proportion of rubble surfaces}
# create new df called 'proportioncover_rubble' that we'll work with:
proportioncover_rubble <- coralnet_prep %>% 
  
# group the existing data so that future operations occur by container and rubble_type  
  group_by(container, rubble_type) %>% 

# create columns with slightly more useful names that count how many times each label was used to annotate a point on each rubble piece  
  summarize(bare_bot = sum(label == "bb"),
            cca_bot = sum(label == "cb"),
            ram_bot = sum(label == "rb"),
            spon_bot = sum(label == "ob"),
            bare_side = sum(label == "bs"),
            cca_side = sum(label == "cs"),
            ram_side = sum(label == "rs"),
            spon_side = sum(label == "os"),
            bare_top = sum(label == "bt"),
            cca_top = sum(label == "ct"),
            ram_top = sum(label == "rt"),
            spon_top = sum(label == "ot")) %>% 
  
# create 5 new columns that count top, side, bottom, CCA, BARE, RAM, and OTHER annotations. Also calculate a total number of annotations per rubble. 
  mutate(count_cca = cca_bot + cca_side + cca_top,
         count_ram = ram_bot + ram_side + ram_top,
         count_bare = bare_bot + bare_side + bare_top,
         count_spon = spon_bot + spon_side + spon_top,
         count_total = count_cca + count_ram + count_bare + count_spon) %>%
  
# create 7 new columns that give the proportion of all annotations on each rubble that are considered top, side, bottom, RAM, BARE, CCA, and OTHER.    
    mutate(prop_cca = count_cca / count_total,
           prop_ram = count_ram / count_total,
           prop_bare = count_bare / count_total,
           prop_spon = count_spon / count_total)



# if you wanted to do the same thing and be able to calculate proportion of each rubble that was top/side/bottom, insert these into the appropriate mutate functions.
        
   #      count_bot = bare_bot + cca_bot + ram_bot + spon_bot,
   #      count_side = bare_side + cca_side + ram_side + spon_side,
   #      count_top = bare_top + cca_top + ram_top + spon_top,

   #      prop_bot = count_bot / count_total,
   #      prop_side = count_side / count_total,
   #      prop_top = count_top / count_total,
```


3. Do the exact same steps but only group by container. This gives patch-level data because all surfaces are considered regardless of what the rubble type was. Top/side/bottom data is left out because we don't need it twice. 
```{r proportion of patch surfaces}
# create new df called 'proportioncover_patch' that we'll work with:
proportioncover_patch <- coralnet_prep %>% 
  
# group the existing data so that future operations occur by container and rubble_type  
  group_by(container) %>% 

# create columns with slightly more useful names that count how many times each label was used to annotate a point in each container.  
  summarize(bare_bot = sum(label == "bb"),
            cca_bot = sum(label == "cb"),
            ram_bot = sum(label == "rb"),
            spon_bot = sum(label == "ob"),
            bare_side = sum(label == "bs"),
            cca_side = sum(label == "cs"),
            ram_side = sum(label == "rs"),
            spon_side = sum(label == "os"),
            bare_top = sum(label == "bt"),
            cca_top = sum(label == "ct"),
            ram_top = sum(label == "rt"),
            spon_top = sum(label == "ot")) %>% 
  
# create 5 new columns that count CCA, BARE, RAM, and OTHER annotations. Also calculate a total number of annotations per container. 
  mutate(count_cca = cca_bot + cca_side + cca_top,
         count_ram = ram_bot + ram_side + ram_top,
         count_bare = bare_bot + bare_side + bare_top,
         count_spon = spon_bot + spon_side + spon_top,
         count_total = count_cca + count_ram + count_bare + count_spon) %>%
  
# create 4 new columns that give the proportion of all annotations in each container that are considered RAM, BARE, CCA, and OTHER.    
    mutate(prop_cca = count_cca / count_total,
           prop_ram = count_ram / count_total,
           prop_bare = count_bare / count_total,
           prop_spon = count_spon / count_total)

proportioncover_patch
```




------------------- PART 2: SIZE DATA FOR CALCULATING AREAS OF CONTAINERS AND RUBBLE PIECES -----------------------

This section includes two different ways of calculating surface area of each rubble piece. The outputs using these methods are named 'method1' and 'method2' to avoid confusion later on, when the word 'area' is used a lot. 

4. Area of each side: This uses data from the imported size_prep to calculate the area and perimeter of the top, area and perimeter of the bottom, area of the sides, and total area of each rubble. This version combines 3 different area and perimeter measurements. This chunk contains "method 1" of area calculation. 
```{r area of each side}
# create a new df called method_1 that calculates total surface areas based on an average of 3 sets of measurements
method1 <- size_prep %>%
  
# add columns that give one average measurement each for the bottom area, top area, bottom perimeter, top perimeter, and height. 
  mutate(area_bot = rowMeans(select(., area_bot1, area_bot2, area_bot3), na.rm = TRUE), 
         area_top = rowMeans(select(., area_top1, area_top2, area_top3), na.rm = TRUE), 
         per_bot = rowMeans(select(., perim_bot1, perim_bot2, perim_bot3), na.rm = TRUE), 
         per_top = rowMeans(select(., perim_top1, perim_top2, perim_top3), na.rm = TRUE),
         height = rowMeans(select(., hgt1, hgt2, hgt3, hgt4), na.rm = TRUE)) %>% 
  
# add columns that calculate the area of the side (average height * average of top & perimeter) and add it to the averaged areas for top and bottom of each rubble piece. 
  mutate(area_side = height * rowMeans(select(., per_bot, per_top)),
         area_wholerubble = area_bot + area_top + area_side) %>%
  
# remove the extra columns that contain the raw measurements so that only the area_bot, area_top, area_side, and area_total remain
  select(-per_bot, -per_top, -height, -ends_with(c("1", "2", "3", "4"))) 

method1
```


5. The chunk above works, but I think that the better approach here is to only use the second set of measurements (area_bot2, area_top2, perim_bot2, perim_top2) because that set was done by tracing the entire rubble perimeter, which includes side measurements. I think the best way to avoid overinflation of the side area calculations is to just take areas and perimeters that include the most area and use the CoralNet annotations to estimate the proportion of the rubble that falls into each side category. This way we are extrapolating as little as possible, whereas the other method requires taking estimates and applying a percent cover to a derived 'side' area. Even if this set of measurements is replaced later, the code should still work. 
This chunk contains "method 2" of area calculation- only data from one set of measurements are included, and CoralNet annotations will be used to estimate side areas instead of the heights. 
The terminology here may be confusing. We took a photo of the "top" and "bottom" side of each rubble piece, but almost all of the side area was included in almost every photo because the rubble pieces were slightly rounded. We will consider all area visible in both photos to be the entire area of the rubble, and will apply CoralNet annotations to estimate how much area is top/side/bottom.  
Although, if we aren't analyzing by top/side/bottom, this is unneccessary and calculations should only require the sum of area_top and area_bot. I think this may reasonable because the rubble pieces were almost entirely visible between the two photos and not so rugose that flattening them for the sake of consistent calculations is going to misconstrue the results more than deriving the area of the sides, which in lots of cases ended up being greater than the sum of top and bottom areas (which is unreasonable inflation). 


```{r}
# create a new df called method2 that calculates total surface areas based on the traced areas of top and bottom to estimate total surface area.
method2 <- size_prep %>%
  
# remove all extra data other than the one set (2) of measurements that we'll use here.   
  select(container, rubble_type, area_bot2, area_top2) %>%

# add a 'total' column with total area of the rubble. 
  mutate(area_wholerubble = area_bot2 + area_top2) %>%

# rename "area_bot2" and "area_top2" to "area_bot" and "area_top" for consistency and ease of switching between the two. Note that this is one measurement and 'method1' area_bot and area_top are averages of 3 rounds of measurements including these. However, those were less precise tracings in order to leave out "sides".
  rename(area_bot = area_bot2,
         area_top = area_top2)

method2
```


6. This is in two parts, with the overall goal of creating a small table that has containers as rows, with 3 colums representing the rubble type. This will show us what proportion of the total SA in each container is contributed by each piece of rubble. We first need to calculate the total area of each container (6a), then insert those values into a column in the dataframe so that every rubble piece is assigned a container number and a total area in that container. From here, we can manipulate the table in a way that results in our goal format as stated previously. 

THIS STEP NEEDS TO BE EDITED IF YOU WANT TO SWITCH BETWEEN AREA CALCULATION METHODS


6a. Create 'container_area' to calculate the total rubble surface area in each container. We're making this so we can add it as a column onto the next step, when we'll need the total area in each container to figure out what proportion of the total each individual rubble piece makes up. 
```{r calculate total available area in each container}
# create 'container_area' which will tell us the total available area in each container (combined total areas from RAM, BARE, and CCA rubble pieces)
container_area2 <- method2 %>%
  
  # only use two of the columns- we aren't interested in top/bottom here but we need to know surface area of each rubble and what container it's in. 
  select(c("container", "area_wholerubble")) %>% 
  
  # group by container and add all 3 rubble areas together. This adds the areas of all the pieces in one container together.  
  group_by(container) %>%
  summarize(area_wholecontainer = sum(area_wholerubble)) 

container_area2
```

6b. Create 'rubble_area2'. This step is for formatting the data in a useful way. We now have the total area in each container and we want to add a column so that each row (rubble piece) contains data about the total area in its container, so that we can divide its area_wholerubble by sum_container_area and determine what proportion of the total area is provided from each piece. 
```{r proportion of the whole container area provided by each rubble piece}
# create 'rubble_area' which will tell us what proportion of the total area in each container each rubble piece is. 
rubble_area2 <- method2 %>%
  
# get rid of 'area_top2' and 'area_bot2' because we don't need them, only totals.    
  select(c("container", "rubble_type", "area_wholerubble")) %>% 
  
# combine our df 'rubble_area2' with 'container_area2'. The argument 'by = "container"' in our merge function tells it to insert the area_wholecontainer value that corresponds to the container number for each row.
  merge(., container_area2, by = "container") %>%

# divide each rubble area by the total container area to get the proportion of each container contributed by each rubble piece (i.e., relative size?).
  mutate(proportion_of_containerArea = area_wholerubble / area_wholecontainer) %>% 

# remove the area_wholerubble and area_wholecontainer columns so we're only left with container, rubble type, and the relative proportion of the whole container area that each rubble represents
  select(-area_wholerubble, -area_wholecontainer) %>%
  
# flip the table so that containers are rows and columns are rubble types
  pivot_wider(names_from = rubble_type, values_from = proportion_of_containerArea, )

rubble_area2
```

A note for part 6:
'method1' can be replaced with 'method2' depending on what the best option is. 
The 'methodX' dataframes contain the surface area of each rubble piece. This replaced a dataframe that chris had called 'area' that is renamed as 'method1' because I wanted to try multiple ways of calculating areas.
'container_areaX' and 'rubble_areaX' have a number that corresponds to the 'methodX' that was used to generate them. These will need to be changed at the end when the area method is determined, but here I have used 2 because I think it's more realistic. 


7. Calculating available area for each type of substratum and patch. The goal here is to end up with a table like the one from part 6, but with 
We're starting with proportioncover_rubble, which contains all of the annotations data after manipulation, including counts for each raw annotation, summed counts per category (top/bot/side, cca/ram/bare/spon), and proportions of each rubble represented by each category. We will merge this with 'method2' (top, bottom, and total areas for each rubble piece) to calculate actual areas. 

THIS CHUNK IS FOR METHOD 2 ONLY! USE THE NEXT ONE FOR METHOD 1
```{r patch type proportions - METHOD2}

# check out what we're starting with and what data area available
head(proportioncover_rubble)
head(method2)

# create new dataframe 'patch_type_area'. We will start with the proportioncover_rubble dataframe (all annotations data) and merge with 'method2'. 
patch_type_area2 <- proportioncover_rubble %>%

# merge the dataframes. This just adds the 3 columns unique to method2 (area_bot, area_top, area_wholerubble) to the end.      
  merge(., method2, by = c("container", "rubble_type")) %>% 

# this cuts out a lot of extra columns. Removes total counts (e.g., count_ram) and proportions of the total, and only leaves the annotation categories and total count. 
  select(container, rubble_type, prop_cca, prop_ram, prop_bare, prop_spon, area_wholerubble) %>% 
  
# calculate the actual area of each type of substratum on each rubble piece by multiplying each proportion of the area by the total area of the rubble. 
  mutate(area_bare = prop_bare * area_wholerubble,
         area_cca = prop_cca * area_wholerubble,
         area_ram = prop_ram* area_wholerubble,
         area_spon = prop_spon * area_wholerubble) %>% 

# group by container so we can get a combined amount of each type between the 3 rubble pieces in each, then create a column that adds up all of the areas for each substratum type. Now we have area per each rubble and we combined each rubble to get total area of each substratum per container. 
  group_by(container) %>%
  summarize(bare = sum(area_bare),
            cca = sum(area_cca),
            ram = sum(area_ram),
            spon = sum(area_spon),
            total = sum(area_wholerubble)) %>% 
  
# replace total areas with proportions of the total: divide individual substratum areas by the total area in the container to find the patch-level surface area proportions. We're renaming the same columns because we want proportion. 
  mutate(bare = bare / total,
         cca = cca / total,
         ram = ram / total,
         spon = spon / total) %>% 
  
# remove total column because we don't need it anymore
  select(-total) %>% 
  
# replace 0's with extremely small number for electivity calculations because we can't have 0's in the data.  
  mutate(bare = replace(bare, bare == 0, 0.000000001),
         cca = replace(cca, cca == 0, 0.000000001),
         ram = replace(ram, ram == 0, 0.000000001),
         spon = replace(spon, spon == 0, 0.000000001)) 

patch_type_area2
```



THIS CHUNK IS FOR METHOD 1 ONLY! USE THE PREVIOUS ONE FOR METHOD2!!!! They have slightly different structures and need to be handled separately here. 
```{r patch type proportions - METHOD1 ONLY}

# check out what we're starting with and what data area available
head(proportioncover_rubble)
head(method1)

# create new dataframe 'patch_type_area'. We will start with the proportioncover_rubble dataframe (all annotations data) and merge with 'method2'. 
patch_type_area1 <- proportioncover_rubble %>%

# merge the dataframes. This just adds the 3 columns unique to method2 (area_bot, area_top, area_wholerubble) to the end.      
  merge(., method1, by = c("container", "rubble_type")) %>% 

# remove extra columns and only keep what we'll need for the rest
  select(container, rubble_type, prop_cca, prop_ram, prop_bare, prop_spon, area_wholerubble) %>% 
  
# calculate the area of each annotation type  
  mutate(area_bare = prop_bare * area_wholerubble,
         area_cca = prop_cca * area_wholerubble,
         area_ram = prop_ram* area_wholerubble,
         area_spon = prop_spon * area_wholerubble) %>% 

# group by container so we can get a combined amount of each type between the 3 rubble pieces in each, then create a column that adds up all of the areas for each substratum type. Now we have area per each rubble and we combined each rubble to get total area of each substratum per container. 
  group_by(container) %>%
  summarize(bare = sum(area_bare),
            cca = sum(area_cca),
            ram = sum(area_ram),
            spon = sum(area_spon),
            total = sum(area_wholerubble)) %>% 

# replace total areas with proportions of the total: divide individual substratum areas by the total area in the container to find the patch-level surface area proportions. We're renaming the same columns because we want proportion. This replaces the 'across()' function. 
  mutate(bare = bare / total,
         cca = cca / total,
         ram = ram / total,
         spon = spon / total) %>% 
 
# remove total column because we don't need it anymore     
  select(-total) %>% 

# replace 0's with extremely small number for electivity calculations because we can't have 0's in the data.    
  mutate(bare = replace(bare, bare == 0, 0.000000001),
         cca = replace(cca, cca == 0, 0.000000001),
         ram = replace(ram, ram == 0, 0.000000001),
         spon = replace(spon, spon == 0, 0.000000001)) 

patch_type_area1
```

8. Export data so we can use it in the electivity/mortality Rmd! 
```{r export}
write_csv(rubble_area2, "rubble_area2.csv") #export the rubble area df
write_csv(patch_type_area2, "patch_type_area2.csv") #export the patch type area df
```









#### CHRIS STUFF UNDER HERE DONT DELETE but everything works and if I used the same data I bet I would get the same answers so I think it's ok 


3. Area of each side: 
```{r area of each side}
area <- size_prep %>%
  mutate(area_bot = rowMeans(select(., area_bot1, area_bot2, area_bot3), na.rm = TRUE), #bottom area
         area_top = rowMeans(select(., area_top1, area_top2, area_top3), na.rm = TRUE), #top area
         per_bot = rowMeans(select(., perim_bot1, perim_bot2, perim_bot3), na.rm = TRUE), #perim of bottom
         per_top = rowMeans(select(., perim_top1, perim_top2, perim_top3), na.rm = TRUE), #perim of top
         height = rowMeans(select(., hgt1, hgt2, hgt3, hgt4), na.rm = TRUE)) %>% #height
  mutate(area_side = height * rowMeans(select(., per_bot, per_top)), #side area
         area_total = area_bot + area_top + area_side) %>% #total area
  select(-per_bot, -per_top, -height, -ends_with(c("1", "2", "3", "4"))) #remove extra columns

area
```

```{r area of each rubble}
container_area <- area %>%
  select(c("container", "area_total")) %>% #only use two of the columns
  group_by(container) %>%
  summarize(area = sum(area_total)) #group by container and add all areas together 
rubble_area <- area %>%
  select(c("container", "rubble_type", "area_total")) %>% #only keep a few columns
  merge(., container_area, by = "container") %>% #give a total area for each container
  mutate(area_total = area_total / area) %>% #and then divide each rubble area to get proportion
  select(-area) %>% #remove the total area column
  pivot_wider(names_from = rubble_type, values_from = area_total, )
```

```{r area of each patch type}
patch_type_area <- prop %>%
  merge(., area, by = c("container", "rubble_type")) %>% #merge the dataframes
  select(ends_with(c("container", "type", "bot", "top", "side", "total"))) %>% #cut out a lot of columns
  mutate(bare_bot = bare_bot * area_bot,
         cca_bot = cca_bot * area_bot,
         ram_bot = ram_bot * area_bot,
         spon_bot = spon_bot * area_bot,
         bare_side = bare_side * area_side,
         cca_side = cca_side * area_side,
         ram_side = ram_side * area_side,
         spon_side = spon_side * area_side,
         bare_top = bare_top * area_top,
         cca_top = cca_top * area_top,
         ram_top = ram_top * area_top,
         spon_top = spon_top * area_top) %>% #find area of each substratum type
  mutate(bare = bare_bot + bare_side + bare_top,
         cca = cca_bot + cca_side + cca_top,
         ram = ram_bot + ram_side + ram_top,
         spon = spon_bot + spon_side + spon_top) %>% #find total area for each type
  group_by(container) %>%
  summarize(bare = sum(bare),
            cca = sum(cca),
            ram = sum(ram),
            spon = sum(spon),
            total = sum(area_total)) %>% #add rubble together
  mutate(across(c(2:5), .fns = ~./total)) %>% #divide areas by total
  select(-total) %>% #remove total column
  mutate(bare = replace(bare, bare == 0, 0.000000001),
         cca = replace(cca, cca == 0, 0.000000001),
         ram = replace(ram, ram == 0, 0.000000001),
         spon = replace(spon, spon == 0, 0.000000001)) #replace 0's with extremely small number for electivity calculations

```


```{r export}
write_csv(rubble_area, "rubble_area.csv") #export the rubble area df
write_csv(patch_type_area, "patch_type_area.csv") #export the patch type area df
```